---
output: github_document

---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "Figures/Lifespan_v1.0.0-",
                      out.width = "100%",
                      cache = TRUE,
                      cache.lazy = FALSE)
# for tibbles...
options(pillar.neg=F, # do no print neg number in red
        pillar.subtle=F, # turn off highliting of significant digits
        tibble.width = 130) # default=95, increase it to make it readable
```


# Lifespan Analysis

In this example, we will use the data from Timmers et al to apply our Bayesian GWAS approach to study lifespan.    
Here, we assume that the `bGWAS` package is already installed, that the Z-matrix files have already been downloaded and stored in `"~/ZMatrices"`. If that is not the case, please follow the steps described [here](../README.md). 

```{r download-data, echo=TRUE, message=FALSE, results='hide', cache=F}
library(bGWAS) # bGWAS github version:

# Download data to working directory (~460 MB) if not already here
if(!file.exists("lifegen_phase2_bothpl_alldr_2017_09_18.tsv.gz")) download.file(url = "https://datashare.is.ed.ac.uk/bitstream/handle/10283/3209/lifegen_phase2_bothpl_alldr_2017_09_18.tsv.gz?sequence=1&isAllowed=y", destfile = "lifegen_phase2_bothpl_alldr_2017_09_18.tsv.gz")
```

Now that we have the data in our working directory, we can launch the analysis (with default parameters):

```{r bGWAS, echo=TRUE}
Lifespan_bGWAS = bGWAS(name = "Lifespan_Timmers2019",
                       GWAS = "lifegen_phase2_bothpl_alldr_2017_09_18.tsv.gz")
```



We can now look at the results more in details.

## Prior GWASs used

```{r results1, dpi=300}
coefficients_plot_bGWAS(Lifespan_bGWAS)
```

`r nrow(extract_MRcoeffs_bGWAS(Lifespan_bGWAS))` prior GWASs (risk factors) are used to create the prior, the multivariate causal effect estimates are consistent with what we would expect. On this figure, the multivariate causal effect estimate and the 95\% interval from the multivariate MR model using all chromosomes (black dot and bars) as well as the 22 per-chromosome estimates (grey bars) are represented for each prior GWASs. Coronary Artery Disease (CAD) has the strongest negative effect on lifespan. High Diastolic Blood Pressure (DBP) and Body Mass Index (BMI) also decreases lifespan. We can also see that education, in this case the number of years of schooling, has a positive effect on lifespan.

Overall, the squared correlation between prior and observed effects is about `r round(get_RSquared_bGWAS(Lifespan_bGWAS, "all"), 3)` and goes up to `r round(get_RSquared_bGWAS(Lifespan_bGWAS, "moderate"), 3)` when we consider only SNPs having at least a moderate effect on lifespan (observed p-value < 0.001).     
Using the previous version (Timmers et al), squared correlation was around 0.003 when considering all SNPs and around 0.082 for SNPs having a moderate effect.    

## Results - BF

With this approach, we identified `r nrow(extract_results_bGWAS(Lifespan_bGWAS))` SNPs affecting lifespan through the identified risk factors:

```{r results2, echo=TRUE}
# all hits
knitr::kable(extract_results_bGWAS(Lifespan_bGWAS) %>% mutate(BF = as.character(format(BF, scientific=T, digits=3)), BF_p = as.character(format(BF_p, scientific=T, digits=3))) %>% arrange(chrm_UK10K), digits=3)

# new hits (compared to conventional GWAS)
extract_results_bGWAS(Lifespan_bGWAS) %>%
  mutate(obs_p = 2*pnorm(-abs(z_obs))) %>%
  filter(obs_p>5e-8) %>% 
  arrange(chrm_UK10K) -> New_Hits
knitr::kable(New_Hits %>% mutate(BF = as.character(format(BF, scientific=T, digits=3)), BF_p = as.character(format(BF_p, scientific=T, digits=3))), digits=3)
```

`r nrow(New_Hits)` of them are missed by the conventional GWAS (using same p-value threshold of 5e-8 to call significance).    
Using the previous version (Timmers et al), we identified 7 new variants (using a threshold of 2.5e-8 for both GWAS and bGWAS results). Using a threshold of 5e-8 would have resulted 

```{r results3, dpi=300}
# For the plots, we will use only the new hits
New_Hits %>% 
  transmute(rs=rsid,
		gene = c("CELSR2/PSRC1", "TMEM18", "BCL11A", "ZNF318/ABCC10", "ZC3HC1", "POM21C", "EPHX2/CLU", "BNC2", "GADD45G", "PDE3A", "LSM4/PGPEP1"),
		color="#932735") -> my_SNPs

manhattan_plot_bGWAS(Lifespan_bGWAS, SNPs=my_SNPs)
```


```{r results4, dpi=300}
my_SNPs %>%
  mutate(color=NULL) -> my_SNPs
heatmap_bGWAS(Lifespan_bGWAS, SNPs = my_SNPs)
```

    

On this figure, the contribution of each risk factor to the prior effects of new hits (alleles aligned to be life-lengthening) is represented as a heatmap. Overall, we observe a lot of red, as expected since alleles are aligned to be life-lengthening.    
Among these 11 new variants, 4 were known to be associated with at least one of the RFs (indicated with a star on the heatmap - variant near CELSR2/PSRC1 associated with LDL cholesterol, variants near TMEM18 and LSM4/PGPEP1 associated with Body Mass Index, variant near BCL11A associated with Years of Schooling). 
7 variants (near ZNF318, ZC3HC1, POM21C, EPHX2/CLU, BNC2, GADD45G and PDE3A) are not associated with any of the RFs (at least not in the summary statistics used to create the prior), suggesting that they could be acting on lifespan through smaller pleiotropic effects on several RFs.    
These variants can be further investigated using the [GWAS Catalog](https://www.ebi.ac.uk/gwas/) and looking at SNPs in 500kb region, using [LDlink](https://ldlink.nci.nih.gov/) to get R2 estimates in EUR population.    
**ZNF318**:      
- a variant in LD (R2=0.8401), rs2270860, associated with "Diastolic blood pressure (cigarette smoking interaction)" in [this paper](https://www.ncbi.nlm.nih.gov/pubmed/26390057)      
- a variant in LD (R2=0.8978), rs16896398, associated with "DBP" in a recent GWAS using Japanese population [here](https://www.ncbi.nlm.nih.gov/pubmed/29403010)      
**ZC3HC1**     
- a variant in LD (R2=0.8661), rs11556924, associated with "CAD" in a recent GWAS using UKBB  [here](https://www.ncbi.nlm.nih.gov/pubmed/29212778)     
**POM21C**          
- two variants in LD (R2=0.1782), rs58862095, associated with "BMI" in a recent GWAS using UKBB  [here](https://www.ncbi.nlm.nih.gov/pubmed/30595370)    
- a variant in LD (R2=0.1947), rs1167827, associated with "Education Attainment" in a recent GWAS [here](https://www.ncbi.nlm.nih.gov/pubmed/30038396) and with "BMI" in [this paper](https://www.ncbi.nlm.nih.gov/pubmed/26426971)    
**EPHX2/CLU**    
- none of the risk factors    
**BNC2**     
- a variant in LD (R2=0.7988), rs10962547, associated with "BMI" in a recent GWAS using UKBB [here](https://www.ncbi.nlm.nih.gov/pubmed/30595370)     
**GADD45G**     
- none of the risk factors     
**PDE3A**    
- none of the risk factors



## Results - Direct Effects


We can use direct effects to identify SNPs significantly acting on lifespan independently from the prior GWASs used to create the prior:

```{r results5, echo=TRUE}
knitr::kable(extract_results_bGWAS(Lifespan_bGWAS, results="direct")  %>% mutate(p_direct=as.character(format(p_direct, scientific=T, digits=3))), digits=3)
```

Here, two variants (rs429358 near APOE and rs8042849 near HYKK/CHRNA3/5) have a significant corrected effect. This is expected since APOE is known to be highly pleiotropic and to notably have an affect on Alzheimer disease, not used to create the prior. The second variant is not associated with any of the risk factor, and therefore its prior effect (0.265) is very small compared to its observed effect (10.659).



\newpage
\Large \textbf{References}
\scriptsize
